Category-based and location-based volitional covert attention affect memory at different timescales

### 2023-03-26 paper/main.tex
minor formatting fixes

\author{Kirsten Ziman$^1, 2$,
Madeline R. Lee$^1$,
Alejandro R. Martinez$^1$,\\
Ethan D. Adner$^1$,
and
Jeremy R. Manning\textsuperscript{$1, \dagger$}\\[0.1in]$^1$Dartmouth College\\
$^2$Princeton University\\
\textsuperscript{$\dagger$}Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

Here we examine the ways different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and place. The stimuli and
presentation durations were constant across the two experiments, but the
experiments differed in how often we asked participants to change the focus of
their attention with respect to image category (face versus place) and image
location (left versus right). After the participants attended to a series of
images, we used a recognition memory test to assess which aspects of the
presented images had been encoded into memory. In both experiments we found
that the images participants covertly attended to were better recognized than
other images, supporting the notion that attention enhances memory
encoding~\citep[i.e., they rated attended images as more familiar than
unattended images][]{Yone02}. After maintaining the focus of attention to a
single image category and location (Sustained Attention Experiment),
participants also recognized the attended-category image at the unattended
location, and (to a lesser extent) the unattended-category image at the
attended location. After more rapidly varying their focus of attention
(Variable Attention Experiment), participants showed a similar boost in
recognition for the unattended-category image at the attended location, but
they did not recognize images at the unattended location. This suggests that
participants were able to shift the location of their covert attentional focus
more rapidly than they were able to shift their focus of covert attention to
stimulus features. We also found differences in the timecourses of these memory
effects, suggesting that the impact of location-based attention on memory
persists on the order of several seconds longer than the impact of
feature-based attention.

>>>>

\author{Kirsten Ziman\textsuperscript{$1, 2$},
Madeline R. Lee$^1$,
Alejandro R. Martinez$^1$,\\
Ethan D. Adner$^1$,
and
Jeremy R. Manning\textsuperscript{$1, \dagger$}\\[0.1in]$^1$Dartmouth College\\
$^2$Princeton University\\
\textsuperscript{$\dagger$}Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

Here we examine the ways different types of volitional covert attention
interact to affect memory. We designed an experimental
paradigm~\citep[following][]{Posn80} that asked participants to attend to a
series of presented composite image pairs while keeping their gaze fixed on a
central point. The image pairs comprised a left and right image, each
constructed by blending an image of a face and place. The stimuli and
presentation durations were constant across the two experiments, but the
experiments differed in how often we asked participants to change the focus of
their attention with respect to image category (face versus place) and image
location (left versus right). After the participants attended to a series of
images, we used a recognition memory test to assess which aspects of the
presented images had been encoded into memory. In both experiments we found
that the images participants covertly attended to were better recognized than
other images, supporting the notion that attention enhances memory
encoding~\citep[i.e., they rated attended images as more familiar than
unattended images;][]{Yone02}. After maintaining the focus of attention to a
single image category and location (Sustained Attention Experiment),
participants also recognized the attended-category image at the unattended
location, and (to a lesser extent) the unattended-category image at the
attended location. After more rapidly varying their focus of attention
(Variable Attention Experiment), participants showed a similar boost in
recognition for the unattended-category image at the attended location, but
they did not recognize images at the unattended location. This suggests that
participants were able to shift the location of their covert attentional focus
more rapidly than they were able to shift their focus of covert attention to
stimulus features. We also found differences in the timecourses of these memory
effects, suggesting that the impact of location-based attention on memory
persists on the order of several seconds longer than the impact of
feature-based attention.
